---
title: "datap Specification"
author: "Christoph Glur"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    toc_depth: 5
  pdf_document:
    toc: yes
    toc_depth: '5'
vignette: |
  %\VignetteIndexEntry{datap Specification} %\VignetteEngine{knitr::rmarkdown} \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> “If I could do it all again, I'd be a plumber.”
>
> -- <cite>Albert Einstein</cite>

# Preliminary Remarks

## Document Version

- datap version: 0.1
- Document Version: 0.1
- Date: 2016-04-21
- License:

## Scope

datap is a lightweight DSL (Domain Specific Language) to define configurable, modular, and re-usable data processes in the R programming language. datap contexts can be used to acquire, pre-process, quality-assure, and merge data.

In practice, each datap setup will consist of the following elements:

1. One or more **datap contexts**. Each context is defined in in a yaml file and contains a series of **taps**. Each tap represents 
a specific datset, as well as their source and pre-processing steps.
2. One or more **R packages**, or your own R scripts contianing functions. The R functions do the actual units of work of the pre-processing steps defined in (1), like e.g. downloading data from the internet, data cleaning, merging, etc. The packages are typically datap-agnostic.
3. The **datap interpreter**, i.e. the R datap package. The interpreter parses the datap context, and maps pre-processing steps defined in (1) to actual library functions available in (2).

This document is about the first part only: the datap context file definition.

## Syntax Description Conventions

In this document, the datap syntax is described using the following conventions:

* `>`: a reference to a specific datap element
* `[]`: optional elements
* `@`: replace the following string with an appropriate name
* `n*`: repeat the element n times
* `|`: or

# datap Syntax

## *context*

A datap [`>context`](#context) is defined in a single YAML document. A YAML document can contain at most one [`>context`](#context).

A [`>context`](#context) spans a tree whose nodes are each one the following types of *joints*:

* [`>tap`](#tap): entry point to data, can have parameters
* [`>structure`](#structure): organise taps into hierarchies
* flow control:
	* [`>pipe`](#pipe): combine joints serially
	* [`>junction`](#junction): merge multiple sub-joints into one
* [`>processor`](#processor): unit of work (data acquisition and pre-processing)
* error handling:
	* [`>warning`](#warning)
	* [`>error`](#error)

Each joint consists of the following:

1. a mandatory *type* (tap, structure, pipe, junction, processor, warning, error)
2. depending on the type, named elements, namely;
    + [`>attributes`](#attributes)
    + [`>variables`](#variables)
    + [`>parameters`](#parameters)
    + [`>function`](#function)
    + [`>condition`](#condition)
3. other, nested *joints*

The flow of data is from leafs towards the root, and ends at a [`>tap`](#tap). Thus, each sub-tree below a [`>tap`](#tap) defines the processing steps of a [`>tap`](#tap). In line with data flow, we use the term *upstream* to denote joints that are processed before a given joint. We use *downstream* to denote joints that are processed after a given joint.

## *variables*

Variables can be defined in a [`>structure`](#structure), [`>tap`](#tap), [`>pipe`](#pipe), and [`>junction`](#junction).

A [`>variables`](#variables) section is an *associative list*, called "variables". Each variable is an entry in that list, with the *key* defining the variable *name*, and the *value* defining the variable *value*:

```
>structure|>tap|>pipe|>junction
  variables:
    n* $variableName: $value
```

The names of *[special references](#special-references)* cannot be used as variable name (namely: "inflow", "joint", "context").

The *scope* of a variable is the sub-tree spanned by the joint in which the variable is defined. A variable value can be overwritten by an upstream joint.

Example:

```{YAML}
Closing Prices:
  type: structure
  variables:
    series: Close
    startDate: 2000-01-01
```


## *reference*

### *variable* and *parameter* reference

A [`>reference`](#reference) has an *`$`* prefix, and refers to a downstream [`>variable`](#variables), a [`>parameter`](#parameters), or a [`>special  reference`](#special-reference).

You can use a [`>reference`](#reference) in a [`>parameter`](#parameters), a [`>function`](#function), or in another [`>variable`](#variables).

```
>parameters|>function|>variables:
  $name: $@variableReferenceName
```

For example:

```{YAML}
AAPL:
  type: tap
  variables:
    #variable reference
    #maxNaRatioDefault must be defined upstream
    maxNaRatio: $maxNaRatioDefault
```

### special reference

The following variable references can be used without defining the variables downstream:

* [`$inflow`](#$inflow)
* [`$joint`](#$joint)
* [`$context`](#context)

They are *reserved words* and cannot be used as variable names.

#### *`$inflow`*

The *`$inflow`* reference refers to the output of the upstream joints. For a *pipe*, there is a single upstream joint. For a *junction*, there can be more than one. In that case, the *`$inflow`* refers to the set of upstream outputs.

Example:

```{YAML}
MinLength:
  type: error
  function: MinLength
  arguments:
    timeseries: '$inflow'
    minLength: 10
```

#### *`$joint`*

The *`$joint`* reference refers to the upstream joints. This is particularly useful in connection with *[factory](#factory)* joints.

Example:

```{YAML}
Cache:
  type: factory
  function: Cache
  arguments:
    f: '$joint'
    timeout: 3600
```

#### *`$context`*

The *`$context`* reference refers to its surrounding [`>context`](#context).

It is useful to source data from within a [`>context`](#context), and to re-use it as an input into another [`>tap`](#tap).

For example:

```{YAML}
Tap:
  type: processor
  function: Tap
  arguments:
    context: '$context'
    tapPath: 'Closing Prices/Indices/SPX'
```


## *attributes*

Attributes can contain information and/or meta data that is not part of the datap processing. For example, for each data series you can store a long name, description, etc. The datap interpreter then provide additional functionality, e.g. to find a [`>tap`](#tap) by attribute.

```
>pipe|>junction|>processor|>factory|>warning|>error|>structure
  attributes:
    n* @attributeName: @value
```

## *function*

### Definition

Functions are mapped to normal R functions. 

In case of a [`>processor`](#processor), the function actually does the unit of work and passes on the result to the downstream joint. 

In case of an [`>error`](#error), or [`>warning`](#warning), the function contains the logic to test for the error condition. 

In case of an [`>attribute`](#attribute), [`>variable`](#variable) or [`>parameter`](#parameter) the function 

```
>processor|>error|>warning
  function: @functionName(n* arg1 [= @default1])
```
and

```
>attribute|>variable|>parameter
  @name: @functionName(n* arg1 [= @default1])
```

### Syntax

The function syntax is similar to R, with a few differences:

1. variables can be used, but they must be referenced using '$'. e.g. `sum(2, $param1)`
2. string literals are interpreted as is, without the necessity of quotes, e.g. `paste(a, b)
3. eclipsis (three dots / ...) are not supported

However, you may nest functions, e.g. `sum(2, sum(3, 5))`, or `sum(seq(1, 10))`.
You may also use named parameters, e.g. `sum(2, 3, na.rm = TRUE)`.

### Execution Time

By default, functions are executed on *tap time*, i.e. when a user calls a tap. However, you may provide the
interpreter with a directive to execute the function already at *build time*, i.e. when reading the context file
and creating the context. This is defined by a `.` preceding the function name. 

For example, the following variable takes on a new value each time the downstream tap is called:

```
variables:
  time: Sys.time()
```

To avoid that, you can use the dot-directive:

```
variables:
  time: .Sys.time()
```


However, if the function expression
contains references to elements that are available at tap time only, then the dot-directive is ignored.
For example:

```
MyStructure:
  type: structure
  variables:
    time: Sys.time()
  Substructure:
    type: structure
    variables:
      # $time cannot be resolved at build time, so . is
      # ignored and the execution time is tap
      yesterday: .subtract($time, -24*60*60)
```



## Joints

Joints are the building blocks of any datap context, as explained in the *[Context](#context)* section.

### *structure*

[`>structure`](#structure) joints fulfil two purposes:

* they define a hierarchy of other joints, especially [`>tap`](#tap)
* they provide a scope to [`>variables`](#variables)

In terms of data processing, structures are of no relevance.

```{r, echo=FALSE, comment=""}
library(datap)
s <- datap:::GetSyntaxDefinition()
print(s$structure)
```


> Consequentially:
>
> * a structure may never be upstream from a [`>tap`](#tap)
> * a structure has no other recognizable type declaration than being a named associative list. Thus, any named associative list inside a structure is itself a structure.
> * a pipe may be defined directly on a structure, without a tap. Such a pipe will not be accessible through the context, and its only purpose is to define a re-usable
> module

### *tap*

A [`>tap`](#tap) defines an entry point to specific data, within a context.

>  Conceptually, you can think of a tap as a public function: when you open a tap (think "call the function"), data pours out (think: "data is returned as an output/return value").


```{r, echo=FALSE, comment=""}
print(s$tap)
```


> There are only [`>structure`](#structure) joints downstream from a tap.
> There are no other [`>tap`](#tap) joints upstream from a tap.

#### *parameters*

A [`>tap`](#tap) may have 0 to n [`>parameters`](#parameters), allowing the caller of the tap to provide tap arguments.

By default, when you define a [`>tap`](#tap) in a [`>context`](#context), all non-resolved upstream function parameters are added as [`>parameter`](#parameters) to the  [`>tap`](#tap). 

However, you may also wish to define [`>parameters`](#parameters) explicitly, mainly for the following reasons:

* to define the order of the parameters
* to provide default arguments for parameters 

```
>tap
  parameters:
    n* $parameterName: [$defaultArgument]
```

For example:

```{YAML}
AAPL: #tap name
  type: tap
  parameters:
    startDate: 2000-01-01
    endDate: Sys.Date()
    includeWeekends:
```

### *processor*

Processors are the work-horse of a [`>tap`](#tap). Each [`>processor`](#processor) defines a unit of work, such as data acquisition, cleaning, or other forms of pre-processing. 

```{r, echo=FALSE, comment=""}
print(s$tap)
```


### *error* and *warning*

[`>error`](#error) and [`>warning`](#warning) joints allow testing the results of the upstream [`>processor`](#processor) joint.

[`>error`](#error) and [`>warning`](#warning) joints are pass-through: the downstream $inflow and $joint variable references the joint's upstream joint.

An [`>error`](#error) condition is a directive to the interpreter to stop execution and display an error message.
A [`>warning`](#warning) condition is a directive to continue execution, and display a warning message.

```{r, echo=FALSE, comment=""}
print(s$error)
```

```{r, echo=FALSE, comment=""}
print(s$warning)
```


Example:

```{YAML}
MinLength:
  type: error
  function: MinLength
  arguments:
    timeseries: '$inflow'
    minLength: 10
```


### *factory*

[`>factory`](#factory) adds functional programming elements to datap.

A [`>factory`](#factory) is similar to a [`>processor`](#processor). The difference is that:

1. a factory's [`>function`](#function) is executed only once, at [`>context`](#context) **creation time** (and not at [`>tap`](#tap) **call time**)
2. the result of the [`>function`](#function) is expected to be itself a [`>function`](#function). That [`>function`](#function) will then be invoked at [`>tap`](#tap) call time.

```{r, echo=FALSE, comment=""}
print(s$factory)
```

Example:

```{YAML}
Cache:
  type: factory
  function: Cache
  arguments:
    f: '$joint'
    timeout: 3600
```

> Interpretation: The *function* of the upstream joint is passed into the Cache function as its *f* argument. *Cache* is expected to be a function factory that returns, as
> an output a memoised version of $joint.


### *pipe*

A [`>pipe`](#pipe) joint lets you arrange a number of upstream joints sequentially.

```{r, echo=FALSE, comment=""}
print(s$pipe)
```


For example, the following [`>pipe`](#pipe) first checks if the number of NAs in a series is below an inacceptable threshold (*NA Ratio*), then it backfills missing values (*Fill NAs*):

```{YAML}
NA handling: &NaHandling
  type: pipe
  Fill NAs:
    type: processor
    function: zoo::na.locf
    arguments:
      object: '$inflow'
  NA Ratio:
    type: warning
    function: NaRatio
    arguments:
      timeseries: '$inflow'
      variable: '@series'
      maxRatio: '@maxNaRatio'
```

### *junction*

A [`>junction`](#junction) merges multiple upstream joints into a single stream.

Unlike the [`>pipe`](#pipe), the [`>junction`](#junction) has a [`>function`](#function), which is a directive how to merge the upstream joints.

```{r, echo=FALSE, comment=""}
print(s$junction)
```


### *module*

Modularization is achieved with YAML *anchors* and *references*. Modules that are not used in a tap can be put in a module section.

```{r, echo=FALSE, comment=""}
print(s$module)
```


For example:

```
modules:
  type: module
  #this module has no tap
  #it only serves as anchors for other taps
  NA handling: &NaHandling
    type: pipe
    Fill NAs:
      type: processor
      function: zoo::na.locf
      arguments:
        object: '$inflow'
    NA Ratio:
      type: warning
      function: NaRatio
      arguments:
        timeseries: '$inflow'
        variable: '@series'
        maxRatio: '@maxNaRatio'

```

# Example

```{r, echo = FALSE, comment = ""}
filePath <- system.file("extdata", "context1.yaml", package="datap")
yamlString <- paste0(readLines(filePath), collapse = "\n")
cat(yamlString)

```
