---
title: "datap Specification"
author: "Christoph Glur"
date: '`r Sys.Date()`'
output:
  html_document:
    toc: yes
    toc_depth: 5
  pdf_document:
    toc: yes
    toc_depth: '5'
vignette: |
  %\VignetteIndexEntry{datap Specification} %\VignetteEngine{knitr::rmarkdown} \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE,
  tidy = FALSE,
  cache = FALSE
)

options("getSymbols.warning4.0"=FALSE)
```


> “If I could do it all again, I'd be a plumber.”
>
> -- <cite>Albert Einstein</cite>

# Preliminary Remarks

## Document Version

- datap version: 0.1
- Document Version: 0.1
- Date: 2016-04-21
- License:

## Introduction

datap is a lightweight DSL (Domain Specific Language) to define configurable, modular, and re-usable data processes for use in the R programming language. use datap contexts to acquire, pre-process, quality-assure, merge and cache any data, of any shape, from any source. Organise your data hierarchically, and access it with a simple, standard api. 

In practice, each datap setup will consist of the following elements:

1. A **datap context**
2. R functions (from R packages, or also your own)
3. The **datap interpreter**, i.e. te R datap package
4. Your R code with which you access your data, by calling API function on the datap interpreter

Event though this document is about the first part mainly (the datap context file definition), let us briefly illustrate what is behind each of the three elements.

### datap context

Each context is defined in in a yaml file and contains a series of hierarchically organised **taps**. Each tap represents a specific dataset. Below a tap, you define the data source and pre-processing steps.

Consider the following, very simple context that is provided with the datap package as *context2.yaml*:


```{r, echo = FALSE, comment = ""}
filePath <- system.file("extdata", "context2.yaml", package="datap")
yamlString <- paste0(readLines(filePath), collapse = "\n")
cat(yamlString)

```

It defines three taps (Apple, Tesla, and S&P500), and organises stocks and indices neatly in a hierarchical structure.

### R functions

The R functions do the actual units of work of the pre-processing steps defined in (1), like e.g. downloading data from the internet (or from a file, database, or in-memory cache), data cleaning, merging, etc. The packages are typically datap-agnostic.

### datap interpreter

The interpreter parses the datap context, and maps pre-processing steps defined in (1) to actual library functions available in (2). 

Once you have the datap package installed, you can load the context into memory using the `datap::Load` function:

```{r}
library(datap)
filePath <- system.file("extdata", "context2.yaml", package="datap")
context <- Load(filePath)
```

The context can be printed:

```{r}
context
```

### R code to access data

Once your context is loaded into memory, you get - for each tap - an R function that can be called to access a data set. You can then directly navigate to a tap and fetch the data set:

```{r}
teslaBars <- context$stocks$Tesla$tap()
head(teslaBars)
```

For the user of the context, it is completely transparent where the data is coming from and how it is pre-processed. For example, the S&P500 index is downloaded from Yahoo finance and not from Quandl. Yet, the user accesses the dataset in exactly the same way:

```{r}
spx <- context$indices$`S&P500`$tap()
head(spx$GSPC.Open)
```

However, in a real world scenario additional pre-processing steps are necessary if you want to make sure that the structure of the data is also the same for datasets from different sources.

In general, however, your data can be anything, and it can come from any source (the internet, a file, from memory, by calling an R function, or generated on the fly by your context, etc.).

## Syntax Description Conventions

In this document, the datap syntax is described using the following conventions:

* `>`: a reference to a specific datap element
* `[]`: optional elements
* `@`: replace the following string with an appropriate name
* `n*`: repeat the following element n times, where n can be any positive number
* `|`: or

# datap Syntax

## *context*

A datap [`>context`](#context) is defined in a single YAML document. A YAML document can contain at most one [`>context`](#context).

A [`>context`](#context) spans a tree whose nodes are each one the following types of *joints*:

* [`>tap`](#tap): entry point to data, can have parameters
* flow control:
	* [`>pipe`](#pipe): combine joints serially
	* [`>junction`](#junction): merge multiple sub-joints into one
* [`>processor`](#processor): unit of work (data acquisition and pre-processing)
* error handling:
	* [`>warning`](#warning)
	* [`>error`](#error)
* [`>module`](#module): define re-usable pipes
* [`>structure`](#structure): organise taps into hierarchies

Each joint consists of the following:

1. a mandatory *type* (tap, structure, pipe, junction, processor, warning, error)
2. named elements, namely (depending on the [`>joint`](#Joints) type);
    + [`>attributes`](#attributes)
    + [`>variables`](#variables)
    + [`>parameters`](#parameters)
    + [`>function`](#function)
    + [`>condition`](#condition)
3. other, nested [`>joints`](#Joints)

The flow of data is from leafs towards the root, and ends at a [`>tap`](#tap). Thus, each sub-tree below a [`>tap`](#tap) defines the processing steps of a [`>tap`](#tap). In line with data flow, we use the term *upstream* to denote joints that are processed before a given joint. We use *downstream* to denote joints that are processed after a given joint.

## *variables*

Variables can be defined in a [`>structure`](#structure), [`>tap`](#tap), [`>pipe`](#pipe), and [`>junction`](#junction).

A [`>variables`](#variables) section is an *associative list*, called "variables". Each variable is an entry in that list, with the *key* defining the variable *name*, and the *value* defining the variable *value*:

```
>structure|>tap|>pipe|>junction
  variables:
    n* $variableName: $value
```

The names of *[special references](#special-reference)* cannot be used as variable name (namely: "inflow", "joint", "context").

The *scope* of a variable is the sub-tree spanned by the joint in which the variable is defined. A variable value can be overwritten by an upstream joint.

Example:

```{YAML}
Closing Prices:
  type: structure
  variables:
    series: Close
    startDate: 2000-01-01
```


## *reference*

### *variable* and *parameter* reference

A [`>reference`](#reference) has an *`$`* prefix, and refers to a downstream [`>variable`](#variables), a [`>parameter`](#parameters), or a [`>special  reference`](#special-reference).

You can use a [`>reference`](#reference) in a [`>parameter`](#parameters) or in a [`>variable`](#variables).

```
>parameters|>variables:
  $name: $@variableReferenceName
```

For example:

```{YAML}
AAPL:
  type: tap
  variables:
    #variable reference
    #maxNaRatioDefault must be defined upstream
    maxNaRatio: $maxNaRatioDefault
```

You can also use a [`>reference`](#reference) in a [`>function`](#function).

For example:

```{YAML}
AAPL:
  type: tap
  variables:
    ticker: "'YAHOO/AAPL'"
  download:
    type: processor
    function: Quandl::Quandl(code = $ticker, type = 'xts')
```

>function|

### special reference

The following variable references can be used without defining the variables downstream:

* [`$inflow`](#$inflow)
* [`$joint`](#$joint)
* [`$context`](#context)

They are *reserved words* and cannot be used as variable names.

#### *`$inflow`*

The *`$inflow`* reference refers to the output of the upstream joints. For a *pipe*, there is a single upstream joint. For a *junction*, there can be more than one. In that case, the *`$inflow`* refers to the set of upstream outputs.

Example:

```{YAML}
MinLength:
  type: error
  function: MinLength
  arguments:
    timeseries: '$inflow'
    minLength: 10
```

#### *`$joint`*

The *`$joint`* reference refers to the upstream joints. This is particularly useful in connection with *[factory](#factory)* joints.

Example:

```{YAML}
Cache:
  type: factory
  function: Cache
  arguments:
    f: '$joint'
    timeout: 3600
```

#### *`$context`*

The *`$context`* reference refers to its surrounding [`>context`](#context).

It is useful to source data from within a [`>context`](#context), and to re-use it as an input into another [`>tap`](#tap).

For example:

```{YAML}
Tap:
  type: processor
  function: Tap
  arguments:
    context: '$context'
    tapPath: 'Closing Prices/Indices/SPX'
```


## *attributes*

Attributes can contain information and/or meta data that is not part of the datap processing. For example, for each data series you can store a long name, description, etc. The datap interpreter then provide additional functionality, e.g. to find a [`>tap`](#tap) by attribute.

```
>pipe|>junction|>processor|>factory|>warning|>error|>structure
  attributes:
    n* @attributeName: @value
```

## *function*

### Definition

Functions are mapped to normal R functions. 

In case of a [`>processor`](#processor), the function actually does the unit of work and passes on the result to the downstream joint. 

In case of an [`>error`](#error), or [`>warning`](#warning), the function contains the logic to test for the error condition. 

In case of an [`>attribute`](#attribute), [`>variable`](#variable) or [`>parameter`](#parameter) the function 

```
>processor|>error|>warning
  function: @functionName(n* arg1 [= @default1])
```
and

```
>attribute|>variable|>parameter
  @name: @functionName(n* arg1 [= @default1])
```

### Syntax

The function syntax is similar to R, with a few differences:

1. datap variables can be used, but they must be referenced using '$'. e.g. `sum(2, $param1)`
2. eclipsis (three dots / ...) are not supported

However, you may nest functions, e.g. `sum(2, sum(3, 5))`, or `sum(seq(1, 10))`.
You may also use named parameters, e.g. `sum(2, 3, na.rm = TRUE)`.

### Package Reference

You can use package notation to refer to a function in a package. For example:

```
Fill NAs:
  type: processor
  function: zoo::na.locf($inflow)
```

### Execution Time

By default, functions are executed on *tap time*, i.e. when a user calls a tap. However, you may provide the
interpreter with a directive to execute the function already at *build time*, i.e. when reading the context file
and creating the context. This is defined by a `.` preceding the function name. 

For example, the following variable takes on a new value each time the downstream tap is called:

```
variables:
  time: Sys.time()
```

To avoid that, you can use the dot-directive:

```
variables:
  time: .Sys.time()
```


However, if the function expression
contains references to elements that are available at tap time only, then the dot-directive is ignored.
For example:

```
MyStructure:
  type: structure
  variables:
    time: Sys.time()
  Substructure:
    type: structure
    variables:
      # $time cannot be resolved at build time, so . is
      # ignored and the execution time is tap
      yesterday: .subtract($time, -24*60*60)
```



## Joints

Joints are the building blocks of any datap context, as explained in the *[Context](#context)* section.

### *structure*

[`>structure`](#structure) joints fulfil two purposes:

* they define a hierarchy of other joints, especially [`>tap`](#tap)
* they provide a scope to [`>variables`](#variables)

In terms of data processing, structures are of no relevance.

```{r, echo=FALSE, comment=""}
library(datap)
s <- datap:::GetSyntaxDefinition()
print(s$structure)
```


> Consequentially:
>
> * a structure may never be upstream from a [`>tap`](#tap)
> * a structure has no other recognizable type declaration than being a named associative list. Thus, any named associative list inside a structure is itself a structure.
> * a pipe may be defined directly on a structure, without a tap. Such a pipe will not be accessible through the context, and its only purpose is to define a re-usable
> module

### *tap*

A [`>tap`](#tap) defines an entry point to specific data, within a context.

>  Conceptually, you can think of a tap as a public function: when you open a tap (think "call the function"), data pours out (think: "data is returned as an output/return value").


```{r, echo=FALSE, comment=""}
print(s$tap)
```


> There are only [`>structure`](#structure) joints downstream from a tap.
> There are no other [`>tap`](#tap) joints upstream from a tap.

#### *parameters*

A [`>tap`](#tap) may have 0 to n [`>parameters`](#parameters), allowing the caller of the tap to provide tap arguments.

By default, when you define a [`>tap`](#tap) in a [`>context`](#context), all non-resolved upstream function parameters are added as [`>parameter`](#parameters) to the  [`>tap`](#tap). 

However, you may also wish to define [`>parameters`](#parameters) explicitly, mainly for the following reasons:

* to define the order of the parameters
* to provide default arguments for parameters 

```
>tap
  parameters:
    n* $parameterName: [$defaultArgument]
```

For example:

```{YAML}
AAPL: #tap name
  type: tap
  parameters:
    startDate: 2000-01-01
    endDate: Sys.Date()
    includeWeekends:
```

### *processor*

Processors are the work-horse of a [`>tap`](#tap). Each [`>processor`](#processor) defines a unit of work, such as data acquisition, cleaning, or other forms of pre-processing. 

```{r, echo=FALSE, comment=""}
print(s$tap)
```


### *error* and *warning*

[`>error`](#error) and [`>warning`](#warning) joints allow testing the results of the upstream [`>processor`](#processor) joint.

[`>error`](#error) and [`>warning`](#warning) joints are pass-through: the downstream $inflow and $joint variable references the joint's upstream joint.

An [`>error`](#error) condition is a directive to the interpreter to stop execution and display an error message.
A [`>warning`](#warning) condition is a directive to continue execution, and display a warning message.

```{r, echo=FALSE, comment=""}
print(s$error)
```

```{r, echo=FALSE, comment=""}
print(s$warning)
```


Example:

```{YAML}
MinLength:
  type: error
  function: MinLength
  arguments:
    timeseries: '$inflow'
    minLength: 10
```


### *factory*

[`>factory`](#factory) adds functional programming elements to datap.

A [`>factory`](#factory) is similar to a [`>processor`](#processor). The difference is that:

1. a factory's [`>function`](#function) is executed only once, at [`>context`](#context) **creation time** (and not at [`>tap`](#tap) **call time**)
2. the result of the [`>function`](#function) is expected to be itself a [`>function`](#function). That [`>function`](#function) will then be invoked at [`>tap`](#tap) call time.

```{r, echo=FALSE, comment=""}
print(s$factory)
```

Example:

```{YAML}
Cache:
  type: factory
  function: Cache
  arguments:
    f: '$joint'
    timeout: 3600
```

> Interpretation: The *function* of the upstream joint is passed into the Cache function as its *f* argument. *Cache* is expected to be a function factory that returns, as
> an output a memoised version of $joint.


### *pipe*

A [`>pipe`](#pipe) joint lets you arrange a number of upstream joints sequentially.

```{r, echo=FALSE, comment=""}
print(s$pipe)
```


For example, the following [`>pipe`](#pipe) first checks if the number of NAs in a series is below an inacceptable threshold (*NA Ratio*), then it backfills missing values (*Fill NAs*):

```{YAML}
NA handling: &NaHandling
  type: pipe
  Fill NAs:
    type: processor
    function: zoo::na.locf
    arguments:
      object: '$inflow'
  NA Ratio:
    type: warning
    function: NaRatio
    arguments:
      timeseries: '$inflow'
      variable: '@series'
      maxRatio: '@maxNaRatio'
```

### *junction*

A [`>junction`](#junction) merges multiple upstream joints into a single stream.

Unlike the [`>pipe`](#pipe), the [`>junction`](#junction) has a [`>function`](#function), which is a directive how to merge the upstream joints.

```{r, echo=FALSE, comment=""}
print(s$junction)
```


### *module*

Modularization is achieved with YAML *anchors* and *references*. Modules that are not used in a tap can be put in a module section.

```{r, echo=FALSE, comment=""}
print(s$module)
```


For example:

```
modules:
  type: module
  #this module has no tap
  #it only serves as anchors for other taps
  NA handling: &NaHandling
    type: pipe
    Fill NAs:
      type: processor
      function: zoo::na.locf
      arguments:
        object: '$inflow'
    NA Ratio:
      type: warning
      function: NaRatio
      arguments:
        timeseries: '$inflow'
        variable: '@series'
        maxRatio: '@maxNaRatio'

```

# Example

```{r, echo = FALSE, comment = ""}
filePath <- system.file("extdata", "context1.yaml", package="datap")
yamlString <- paste0(readLines(filePath), collapse = "\n")
cat(yamlString)

```
